---
title: "WNBA Archetypal Analysis - 2016 - 2018 Regular Seasons"
output: html_notebook
---

## Load data and libraries
```{r}
library(tidyverse)
library(janitor)
WNBA_2016_Player_BoxScore_Dataset <-
  read_excel("~/DFS/NBAStuffer_WNBA/WNBA-2016-Player-BoxScore-Dataset.xlsx") %>% 
  filter(DATASET == "2016 Regular Season") %>% 
  clean_names()
WNBA_2017_Player_BoxScore_Dataset <-
  read_excel("~/DFS/NBAStuffer_WNBA/WNBA-2017-Player-BoxScore-Dataset.xlsx") %>% 
    filter(DATASET == "2017 Regular Season")
colnames(WNBA_2017_Player_BoxScore_Dataset) <-
  colnames(WNBA_2016_Player_BoxScore_Dataset)
WNBA_2018_Player_BoxScore_Dataset <-
  read_excel("~/DFS/NBAStuffer_WNBA/WNBA-2018-Player-BoxScore-Dataset.xlsx") %>% 
  filter(`DATA SET` == "2018 Regular Season")
colnames(WNBA_2018_Player_BoxScore_Dataset) <-
    colnames(WNBA_2016_Player_BoxScore_Dataset)
wnba_data <- bind_rows(
  WNBA_2016_Player_BoxScore_Dataset,
  WNBA_2017_Player_BoxScore_Dataset,
  WNBA_2018_Player_BoxScore_Dataset
)
rm(
  WNBA_2016_Player_BoxScore_Dataset,
  WNBA_2017_Player_BoxScore_Dataset,
  WNBA_2018_Player_BoxScore_Dataset
)
gc(verbose = TRUE)

```

## Augment the raw data
The NBA / WNBA player box score files from BigDataBall have the raw data only. 
For fantasy sports analytics, we need to compute some derived metrics. We need
double-doubles and triple-doubles, and we also compute the FanDuel, DraftKings
and Yahoo fantasy points. Finally, we also compute the "Game Score", another 
popular single-number metric.
```{r}

```

