---
title: "Archetypal Ballers and Ternary Plots"
subtitle: "A look at the 2018 WNBA season"
author: M. Edward (Ed) Borasky <https://znmeb.github.io/dfstools/>
date: June 8, 2019
output: 
  beamer_presentation:
    theme: EastLansing
bibliography: ternary.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

## Load libraries and data
library(tidyverse)
wnba_data <- dfstools::player_box_score(
  "~/DFS/NBAStuffer_WNBA/WNBA-2018-Player-BoxScore-Dataset.xlsx"
)

## Prepare archetypal analysis input
player_totals <- wnba_data %>% 
  mutate(player_name = player_full_name) %>% 
  select(
    player_name, tot, x3p, a, st, to, bl, pts, ddbl, tdbl, min, games) %>% 
  group_by(player_name) %>% 
  summarize_if(is.numeric, sum) %>% 
  ungroup()
player_labels <- player_totals %>% select(player_name)

## Compute the three-archetype model
archetypal_analysis <- dfstools::compute_archetypes(
  player_totals, player_labels, num_archetypes = 3)
player_alphas <- archetypal_analysis[["player_alphas"]] %>% 
  mutate(PlayerCode = as_factor(sprintf(
    "%s: Rin %1.4f, Floor %1.4f",
    player_name, 
    Rim, 
    Floor
  )))

```

## Archetypal Analysis [@Eugster2009]
* Goal:
    - Dimensionality reduction
    - Reduce a dataset with 17 box score metrics to three archetypes
    - Represent players' skills in terms of the archetypes
* Inputs:
    - Player box scores for the 2018 WNBA season
    - About 160 players
* Outputs:
    - Archetype table
    - Player table

## The Three Archetypes
* Weighted box score metrics
* Each archetype represents a skill set
    - Floor Spacing [@Shea2014, chapter 5]: 
        - skills for a backcourt player (Guard) 
        - dominated by three-point shots
    - Rim Protection [@Shea2014, chapter 6]: 
        - skills for a frontcourt player (Center, Forward)
        - dominated by rebounds
    - Bench: the "bad" archetype - players with low overall skills

## The Player Table
* For each player, the percentage of a player's skill in each archetype
* Floor + Rim + Bench = 1
* Rim + Floor = 1 - Bench represents overall skill
* The archetypal player is the one with the largest value for the archetype

## Ternary plots [@Hamilton2018]
* Rim, Floor and Bench sum to 1.0
* So we can plot them on a ternary plot
    - equilateral triangle with the corners labeled Rim, Floor and Bench
    - The poor players cluster near "Bench"
    - The good rebounders cluster near "Rim"
    - The good three-point shooters cluster near "Floor"
    - And the all-stars are near the Rim - Floor line

_Who are the archetypal floor spacers and rim protectors in the WNBA?_


## 2018 WNBA Top Twelve Overall - The All-Stars
```{r}
dfstools::ternary_plot(
  player_alphas %>% 
    top_n(12, -Bench) %>% 
    arrange(Bench) %>% 
    mutate(Player = fct_reorder(PlayerCode, -Rim)), 
  plot_title = "2018 WNBA Top Twelve Overall")

```

## 2018 WNBA Top Twelve Floor Spacers
```{r}
dfstools::ternary_plot(
  player_alphas %>% 
    top_n(12, Floor) %>% 
    arrange(desc(Floor)) %>% 
    mutate(Player = fct_reorder(PlayerCode, -Floor)), 
  plot_title = "2018 WNBA Top Twelve Floor Spacers")

```

## 2018 WNBA Top Twelve Rim Protectors
```{r}
dfstools::ternary_plot(
  player_alphas %>% 
    top_n(12, Rim) %>% 
    arrange(desc(Rim)) %>% 
    mutate(Player = fct_reorder(PlayerCode, -Rim)), 
  plot_title = "2018 WNBA Top Twelve Rim Protectors")

```

## References
